---
title: "Data and group coursework"
subtitle: "SMM048 (2025-26) Group Coursework 01 - Group 07"
author:
  - Ardi Wira Sudarmo
  - Bong Su Kil
  - Yanwun Mak
  - Yeng Issah
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    self_contained: true
 #   css: style.css
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: united
    highlight: tango
    df_print: paged
  bookdown::pdf_document2:
    latex_engine: xelatex
    citation_package: natbib
    number_sections: true
    toc: true
    toc_depth: 3
    keep_tex: true
    includes:
      in_header: preamble.tex
    pandoc_args:
      - "--lua-filter=answer.lua"
#bibliography: references.bib
---
```{=latex}
\newpage
```


```{r setup, include=FALSE}
dir.create("fig", showWarnings = FALSE)

# Defaults common to all outputs
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.path  = "fig/",
  dpi       = 300
)

# Output-specific settings
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "pdf",
    fig.pos = "!ht",
    out.extra = ""
  )
} else {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "svglite"  # or "png"
  )
}
```

```{r setup-qol, include=FALSE, echo=FALSE}
#----------------------------- Clean environment ------------------------------#
rm(list = ls()) # Remove all objects
graphics.off() # Close all graphical devices
cat("\014") # Clean console
```

```{r load-dependencies, include=FALSE, echo=FALSE}
#------------------- Load dependencies / external libraries -------------------#
# List of required packages
packages <- c(
  "quantmod",   # for downloadin
  "lavaan",     # for structural equation modelling (SEM) and confirmatory factor analysis
  "haven",      # to import/export SPSS, Stata, and SAS data files
  "cluster",    # for cluster analysis methods (k-means, hierarchical clustering, etc.)
  "MASS",       # for statistical functions and multivariate analysis
  "xts",        # for extensible time series objects (especially financial time series)
  "lubridate",  # for easy handling and manipulation of dates and times
  "TSA",        # for time series analysis (ARIMA, spectral analysis, etc.)
  "labelled",    # for adding and managing variable labels
  "psych",      #  for exploratory factor analysis (EFA)
  "GPArotation", # for factor rotation methods (e.g., varimax, oblimin, promax)
  "ggplot2",   # for data visualisation using the grammar of graphics
  "insuranceData",   # for actuarial and insurance-related example datasets
  "GJRM", # for copula regressions
  "corrplot", # for correlation hitmap method
  "GGally",
  "reshape2",
  "plotly",
  "dplyr",
  "knitr"
)

# Check which packages are already installed
installed <- packages %in% installed.packages()[,"Package"]

# Install any missing packages
if (any(!installed)) {
  install.packages(packages[!installed], dependencies = TRUE)
}

# Load all packages
lapply(packages, library, character.only = TRUE)
```
# Factor Analysis

An exploratory factor analysis (EFA) was conducted on the 25 personality items measured on a 6-point Likert scale. The dataset contained some missing values (NA), which were handled using listwise deletion. In addition, several negatively worded items (X1, X9, X10, X11, X12, X22, and X25) were reverse-coded so that higher values consistently represented higher levels of the underlying traits. 

```{r Q1_data, echo=knitr::is_html_output(), include=FALSE}
## R Markdown

data1 <- read.table("dataset1.txt", header = TRUE)
sum(is.na(data1))
colSums(is.na(data1)) 
data1_clean <- na.omit(data1)

reverse_items <- c("X1","X9","X10","X11","X12","X22","X25")

data1_clean[reverse_items] <- 7 - data1_clean[reverse_items]

var_label(data1_clean$X1)  <- "Be indifferent to the feelings of others"
var_label(data1_clean$X2)  <- "Inquire about others' well-being"
var_label(data1_clean$X3)  <- "Know how to comfort others"
var_label(data1_clean$X4)  <- "Love children"
var_label(data1_clean$X5)  <- "Make people feel at ease"
var_label(data1_clean$X6)  <- "Be exacting in my work"
var_label(data1_clean$X7)  <- "Continue until everything is perfect"
var_label(data1_clean$X8)  <- "Do things according to a plan"
var_label(data1_clean$X9)  <- "Do things in a half-way manner"
var_label(data1_clean$X10) <- "Waste my time"
var_label(data1_clean$X11) <- "Don't talk a lot"
var_label(data1_clean$X12) <- "Find it difficult to approach others"
var_label(data1_clean$X13) <- "Know how to captivate people"
var_label(data1_clean$X14) <- "Make friends easily"
var_label(data1_clean$X15) <- "Take charge"
var_label(data1_clean$X16) <- "Get angry easily"
var_label(data1_clean$X17) <- "Get irritated easily"
var_label(data1_clean$X18) <- "Have frequent mood swings"
var_label(data1_clean$X19) <- "Often feel blue"
var_label(data1_clean$X20) <- "Panic easily"
var_label(data1_clean$X21) <- "Be full of ideas"
var_label(data1_clean$X22) <- "Avoid difficult reading material"
var_label(data1_clean$X23) <- "Carry the conversation to a higher level"
var_label(data1_clean$X24) <- "Spend time reflecting on things"
var_label(data1_clean$X25) <- "Will not probe deeply into a subject"
```

To determine the appropriate number of factors, the scree plot and parallel analysis were examined. The scree plot shows a sharp decline in eigenvalues for the first few components, with eigenvalues of 5.13, 2.75, 2.14, 1.85, 1.55, and 1.07 for the first six components. According to the Kaiser criterion (eigenvalue > 1), up to six components could be retained. However, the sixth eigenvalue (1.07) is very close to the threshold, and the scree plot begins to level off after the fifth component. Considering the overall results from the scree plot and parallel analysis, \textbf{a five-factor solution} was selected for the subsequent factor analysis.

```{r Q1plot1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Scree plot and parallel analysis for determining the number of factors", fig.show='hold', fig.align='center', fig.width=6, fig.height=3.2, fig.pos='H', out.width='45%'}

# Scree plot
fa.parallel(data1_clean, fa = "fa", main = "Parallel Analysis for Factor Selection")

# PCA
pca1 <- prcomp(data1_clean, scale. = TRUE)
eig <- pca1$sdev^2
pcs <- 1:length(eig)

plot(
  pcs, eig,
  type = "b",
  pch = 19,
  col = "darkgreen",        
  lwd = 2,
  xlab = "Principal Component",
  ylab = "Eigenvalue",
  main = "PCA Scree Plot",
  cex.main = 1.4,
  cex.lab = 1.2,
  cex.axis = 1.1,
  ylim = c(0, max(eig) + 0.3)
)

lines(pcs, eig, col = "red", lwd = 1.5)

idx <- which(eig >= 1)

text(
  pcs[idx] + 0.15,        
  eig[idx],               
  labels = round(eig[idx], 2),
  cex = 0.85,            
  col = "darkblue",
  pos = 4                
)

abline(h = 1, col = "red", lwd = 1.5, lty = 2)

```

Figure \@ref(fig:q1plot2) shows the rotated factor loading matrices obtained using varimax and promax rotations. Both rotations produce broadly similar loading patterns, suggesting that the factor structure is stable. Most items load strongly on a single factor, with limited cross-loadings. The promax rotation allows correlations between factors and yields slightly clearer clustering of items. Overall, the results support the five-factor solution selected from the scree plot and parallel analysis.


``` {r q1plot2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Rotated Factor loadings(Varimax vs Promax)", fig.show='hold', fig.align='center', fig.width=6, fig.height=3.2, fig.pos='H', out.width='45%'}


fa5 <- factanal(
  data1_clean,
  factors  = 5,
  rotation = "varimax",
  scores   = "regression"
)

L  <- as.matrix(fa5$loadings)
df <- as.data.frame(as.table(L))   # Var1=Item, Var2=Factor, Freq=Loading
colnames(df) <- c("Item", "Factor", "Loading")

df$Factor <- factor(df$Factor, levels = colnames(L))
df$Item   <- factor(df$Item,   levels = rev(rownames(L)))  

ggplot(df, aes(x = Factor, y = Item, fill = Loading)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(
    title = "Rotated Factor Loadings (Varimax)",
    x = "Factor",
    y = "Item"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid = element_blank()
  )

fa5_2 <- factanal(
  data1_clean,
  factors  = 5,
  rotation = "promax",
  scores   = "Bartlett"
)

L2  <- as.matrix(fa5_2$loadings)
df2 <- as.data.frame(as.table(L2))   # Var1=Item, Var2=Factor, Freq=Loading
colnames(df2) <- c("Item", "Factor", "Loading")

df2$Factor <- factor(df2$Factor, levels = colnames(L2))
df2$Item   <- factor(df2$Item,   levels = rev(rownames(L2)))  

ggplot(df2, aes(x = Factor, y = Item, fill = Loading)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(
    title = "Rotated Factor Loadings (Promax)",
    x = "Factor",
    y = "Item"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid = element_blank()
  )

```

The uniqueness values range from 0.27 to 0.83, with an average of approximately 0.58, indicating that the extracted factors explain a moderate proportion of variance in the observed items. The rotated factor loadings reveal a clear five-factor structure. Factor 1 loads strongly on items related to emotional instability (X16–X20), suggesting neuroticism. Factor 2 reflects extraversion, with items describing sociability and interpersonal engagement (e.g., X11–X15). Factor 3 corresponds to conscientiousness, capturing organisation, diligence, and careful work behaviour (X6–X10), while Factor 4 represents agreeableness, including empathy and concern for others (X2–X5). Finally, Factor 5 is associated with openness to experience, reflecting intellectual curiosity and reflective thinking (X21–X25).


``` {r q1table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}


df <- df %>%
  filter(abs(Loading) >= 0.40) %>%
  arrange(Factor, desc(abs(Loading)))

factor_scores <- as.data.frame(fa5$scores)
colnames(factor_scores) <- paste0("F", 1:5)

data1_clean_with_scores <- cbind(data1_clean, factor_scores)

uniq <- summary(fa5$uniquenesses)

uniq_df <- as.data.frame(t(uniq))

knitr::kable(
  uniq_df,
  digits = 3,
  align = "c",
  caption = "Summary of Uniqueness Values"
)

```

``` {r q1table2, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}


df$Factor <- gsub("Factor","F",df$Factor)

kable(
  df,
  digits = 2,
  align = "c",
  caption = "Main Factor Loadings"
)
```


```{=latex}
\newpage
```


# copula analysis 

```{r Q2, echo=knitr::is_html_output(), include=TRUE}
# library(GJRM)
# library(insuranceData)

data(dataCar)

data_cop <- subset(dataCar,
                   numclaims > 0 & claimcst0 > 0 & 
			 is.finite(claimcst0) & is.finite(exposure) & 
	   	   	 exposure > 0)

eq1 <- numclaims ~ 1
eq2 <- claimcst0 ~ 1

#claim frequency distribution
#use offset since exposure is not constant
#truncated frequency

tPfreq <- GJRM::gamlss(list(eq1), family = "tP", data = data_cop, offset = data_cop$exposure)
tNBIfreq <- GJRM::gamlss(list(eq1), family = "tNBI", data = data_cop, offset = data_cop$exposure)
tNBIIfreq <- GJRM::gamlss(list(eq1), family = "tNBII", data = data_cop, offset = data_cop$exposure)
tPIGfreq <- GJRM::gamlss(list(eq1), family = "tPIG", data = data_cop, offset = data_cop$exposure) 

print(AIC(tPfreq, tNBIfreq, tNBIIfreq, tPIGfreq))
print(BIC(tPfreq, tNBIfreq, tNBIIfreq, tPIGfreq))
```


```{r}
conv.check(tPfreq)
res.check(tPfreq)
summary(tPfreq)
```


TODO: Marcella put her analysis results here

it's using markdown syntax.

Example latex:
$$
x+1
$$


```{r}

#claim severity distribution

LNsev <- GJRM::gamlss(list(eq2), family="LN", data=data_cop)
IGsev <- GJRM::gamlss(list(eq2), family="IG", data=data_cop)
GAsev <- GJRM::gamlss(list(eq2), family="GA", data=data_cop)

WEIsev <- GJRM::gamlss(list(eq2), family="WEI", data=data_cop)
SMsev <- GJRM::gamlss(list(eq2), family="SM", data=data_cop)
FISKsev <- GJRM::gamlss(list(eq2), family="FISK", data=data_cop)
DAGUMsev <- GJRM::gamlss(list(eq2), family="DAGUM", data=data_cop)

print(AIC(DAGUMsev, LNsev, WEIsev, IGsev, GAsev, SMsev, FISKsev))
print(BIC(DAGUMsev, LNsev, WEIsev, IGsev, GAsev, SMsev, FISKsev))
```


```{r}

conv.check(IGsev)
res.check(IGsev)
summary(IGsev)
```


```{r}

###fitting copula
ZTPIGN <- gjrm(list(eq1, eq2, ~1, ~1), data = data_cop, margin = c("tP", "IG"), model = "B", copula = "N")
ZTPIGT <- gjrm(list(eq1, eq2, ~1, ~1), data = data_cop, margin = c("tP", "IG"), model = "B", copula = "T")
```


```{r}
print(AIC(ZTPIGN, ZTPIGT))
```

```{=latex}
\newpage
```

# principal component Analysis

We applied Principal Component Analysis (PCA), an unsupervised dimensionality-reduction method, to transform the 11 variables into a smaller set of uncorrelated components that successively explain the maximum possible variance. Variables were standardised using z-scores with before PCA because features are in different units, e.g., income vs age vs rating scores. Without scaling, high-variance variables can dominate PC1 purely due to measurement scale rather than the real underlying structure in the data.

```{r Q3 data, echo=knitr::is_html_output(), include=FALSE}
## R Markdown

data3 <- read.table("dataset3.txt", header = TRUE)
data3 <- na.omit(data3)
colnames(data3) <- c(
  "INC",
  "AGE",
  "SPEND",
  "SHOP",
  "LOYAL",
  "WEB",
  "SOCIAL",
  "EMAIL",
  "SAT",
  "PROMO",
  "CART"
)

# PCA with standardisation
pr.out <- prcomp(data3, scale.  = TRUE)

# variance explained
pr.var <- pr.out$sdev^2
pve <- pr.var / sum(pr.var)

```

The correlation matrix shows generally weak to moderate correlations among variables, suggesting that several variables related to purchasing behaviour (spending score, Shopping Frequency, and Brand loyalty) are strongly positively correlated. PC1 contrasts purchasing intensity (spending score, shopping frequency, and brand loyalty) against age, suggesting that older customers tend to have lower purchasing-intensity scores on this component( here can add scatterplot as evidence to support if have space).



```{r Q3Cor, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Correlation matrix (numbers left, heatmap right)", fig.show='hold', fig.align='center', fig.width=6, fig.height=3.2, fig.pos='H', out.width='45%'}
      
corrplot(cor(data3),
         method="number",
         type="upper",
         tl.cex = 0.25, number.cex = 0.65, tl.srt = 45, cl.cex=0.45)

corrplot(cor(data3),
         method="color",
         type="upper",
         col=colorRampPalette(c("blue","white","red"))(200),
         tl.cex = 0.25, number.cex = 0.4, tl.srt = 45, cl.cex=0.45)

```


```{r Q3 PLOT1, eval=knitr::is_html_output(), echo=FALSE}

ggplot(data.frame(PC = 1:length(pve), PVE = pve), aes(x = PC, y = PVE)) +
  geom_line(color = "red", size = 1) +
  geom_point(size = 3, color = "darkgreen") +
  geom_text(
    aes(label = round(PVE, 2)),
    vjust = -0.8,
    color = "black",
    size = 4
  ) +
  ylim(0, 1) +
  labs(title = "Scree Plot with PVE", x = "Principal Component", y = "Proportion of Variance Explained") +
  theme_minimal()

```

TThe cumulative variance curve shows that the first three PCs explain over half of the total variation. Parallel analysis also supports retaining three components, since only the first three observed eigenvalues exceed those from simulated data. (tightened version)

```{r Q3plot2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Cumulative variance, parallel analysis.", fig.show='hold', fig.align='center', fig.width=7, fig.height=3, fig.pos='H', out.width='48%'}      
      
cum_pve <- cumsum(pve)

ggplot(data.frame(PC = 1:length(cum_pve), CUM = cum_pve), aes(x = PC, y = CUM)) +
  geom_line(color = "red", size = 0.6) +
  geom_point(size = 1.6, color = "darkgreen") +
  geom_hline(yintercept = 0.6,
             linetype = "dashed",
             color = "red") +
  geom_text(
    aes(label = round(CUM, 2)),
    vjust = -0.6,
    color = "black",
    size = 2.4
  ) +
  ylim(0, 1.05) +
  labs(title = "Cumulative Variance Explained", x = "Principal Component", y = "Cumulative PVE") +
  theme_minimal(base_size = 8)

op <- par(cex=0.5, mar=c(3,3,2,1))
fa.parallel(data3, fa="pc", main="Parallel analysis", n.iter=100)
par(op)

```
The loading plot suggests that PC1 is driven by Spending Score, Shopping Frequency and Brand Loyalty, with the largest absolute loadings, representing purchasing intensity. Age loads in the opposite direction, indicating lower purchasing intensity among older customers, noting that PC signs are arbitrary. PC2 loads most strongly and positively on Age and Income, with Email Open Rate also positive, so higher PC2 scores reflect relatively older, higher-income customers who are more responsive to email as a communication channel. PC3 has opposite-signed loadings, which are positive for Promotion Usage and Cart Abandonment but negative for Social Media and Email engagement. It separates price-sensitive basket behaviour (promo use, cart abandonment) from digital-channel engagement (social/email). Overall, the dataset can be reduced to three orthogonal behavioural drivers while retaining most of the relevant information.

```{r Q3plot3, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PCA loadings.", fig.show='hold', fig.align='center', fig.width=7, fig.height=3, fig.pos='H', out.width='48%'}      

load <- pr.out$rotation[, 1:3]
df <- melt(load)

ggplot(df, aes(Var2, Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white") +
  labs(title = "Loadings Heatmap (PC1~PC3)",
       x = "Principal Components", y = "Variables") +
  theme_minimal()

load2 <- as.data.frame(pr.out$rotation[,1:2])
load2$var <- rownames(load2)

ggplot(load2, aes(x=PC1, y=PC2, label=var)) +
  geom_hline(yintercept=0, linewidth=0.3) +
  geom_vline(xintercept=0, linewidth=0.3) +
  geom_segment(aes(x=0, y=0, xend=PC1, yend=PC2),
               linewidth=0.4,
               arrow = arrow(length = unit(0.08,"inches"))) +
  geom_text(size=2.4, vjust=-0.4) +
  coord_equal() +
  labs(title="Loadings plot",
       x="PC1 loading", y="PC2 loading") +
  theme_minimal(base_size=8)

```



```{r, echo=knitr::is_html_output(), include=TRUE}

km <- kmeans(data3, centers = 3)
scores <- data.frame(pr.out$x)
scores2 <- data.frame(pr.out$x[,1:2], cluster = factor(km$cluster))
scores3 <- data.frame(pr.out$x[, 1:3])  

round(pr.out$rotation[,1:3], 2)


 ggplot(scores2, aes(PC1, PC2, color = cluster)) +
   geom_point() +
   theme_minimal()

ggplot(df, aes(Var2, Var1, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", high="red", mid="white") +
  labs(title="PCA Loadings", x="PC", y="Variables") +
  theme_minimal(base_size=8)

ggplot(scores, aes(x = PC1, y = PC2)) +
  geom_point(color = "darkgreen", alpha = 0.7) +
  labs(title = "PCA Score Plot (PC1 vs PC2)", x = "PC1", y = "PC2") +
  theme_minimal()



plot_ly(scores3,
        x = ~PC1,
        y = ~PC2,
        z = ~PC3,
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 3, color = scores3$PC1, colorscale = "Viridis")) %>%
  layout(title = "3D PCA Plot (PC1, PC2, PC3)")

```


# Time seires
