---
title: "Data and group coursework"
subtitle: "SMM048 (2025-26) Group Coursework 01 - Group 07"
author:
  - Ardi Wira Sudarmo
  - Bong Su Kil
  - Yanwun Mak
  - Yeng Issah
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::html_document2:
    self_contained: true
 #   css: style.css
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: united
    highlight: tango
    df_print: paged
  bookdown::pdf_document2:
    latex_engine: xelatex
    citation_package: natbib
    number_sections: true
    toc: true
    toc_depth: 3
    keep_tex: true
    includes:
      in_header: preamble.tex
    pandoc_args:
      - "--lua-filter=answer.lua"
#bibliography: references.bib
---
```{=latex}
\newpage
```


```{r setup, include=FALSE}
dir.create("fig", showWarnings = FALSE)

# Defaults common to all outputs
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.align = "center",
  out.width = "100%",
  fig.path  = "fig/",
  dpi       = 300
)

# Output-specific settings
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "pdf",
    fig.pos = "!ht",
    out.extra = ""
  )
} else {
  knitr::opts_chunk$set(
    fig.width = 6,
    fig.height = 4,
    dev = "svglite"  # or "png"
  )
}
```

```{r setup-qol, include=FALSE, echo=FALSE}
#----------------------------- Clean environment ------------------------------#
rm(list = ls()) # Remove all objects
graphics.off() # Close all graphical devices
cat("\014") # Clean console
```

```{r load-dependencies, include=FALSE, echo=FALSE}
#------------------- Load dependencies / external libraries -------------------#
# List of required packages
packages <- c(
  "quantmod",   # for downloadin
  "lavaan",     # for structural equation modelling (SEM) and confirmatory factor analysis
  "haven",      # to import/export SPSS, Stata, and SAS data files
  "cluster",    # for cluster analysis methods (k-means, hierarchical clustering, etc.)
  "MASS",       # for statistical functions and multivariate analysis
  "xts",        # for extensible time series objects (especially financial time series)
  "lubridate",  # for easy handling and manipulation of dates and times
  "TSA",        # for time series analysis (ARIMA, spectral analysis, etc.)
  "labelled",    # for adding and managing variable labels
  "psych",      #  for exploratory factor analysis (EFA)
  "GPArotation", # for factor rotation methods (e.g., varimax, oblimin, promax)
  "ggplot2",   # for data visualisation using the grammar of graphics
  "insuranceData",   # for actuarial and insurance-related example datasets
  "GJRM", # for copula regressions
  "corrplot", # for correlation hitmap method
  "GGally",
  "reshape2",
  "plotly"
)

# Check which packages are already installed
installed <- packages %in% installed.packages()[,"Package"]

# Install any missing packages
if (any(!installed)) {
  install.packages(packages[!installed], dependencies = TRUE)
}

# Load all packages
lapply(packages, library, character.only = TRUE)
```

# Factor Analysis

```{r Q1_data, echo=knitr::is_html_output(), include=TRUE}
## R Markdown

data1 <- read.table("dataset1.txt", header = TRUE)
sum(is.na(data1))
colSums(is.na(data1)) 
data1_clean <- na.omit(data1)

var_label(data1_clean$X1)  <- "Be indifferent to the feelings of others"
var_label(data1_clean$X2)  <- "Inquire about others' well-being"
var_label(data1_clean$X3)  <- "Know how to comfort others"
var_label(data1_clean$X4)  <- "Love children"
var_label(data1_clean$X5)  <- "Make people feel at ease"
var_label(data1_clean$X6)  <- "Be exacting in my work"
var_label(data1_clean$X7)  <- "Continue until everything is perfect"
var_label(data1_clean$X8)  <- "Do things according to a plan"
var_label(data1_clean$X9)  <- "Do things in a half-way manner"
var_label(data1_clean$X10) <- "Waste my time"
var_label(data1_clean$X11) <- "Don't talk a lot"
var_label(data1_clean$X12) <- "Find it difficult to approach others"
var_label(data1_clean$X13) <- "Know how to captivate people"
var_label(data1_clean$X14) <- "Make friends easily"
var_label(data1_clean$X15) <- "Take charge"
var_label(data1_clean$X16) <- "Get angry easily"
var_label(data1_clean$X17) <- "Get irritated easily"
var_label(data1_clean$X18) <- "Have frequent mood swings"
var_label(data1_clean$X19) <- "Often feel blue"
var_label(data1_clean$X20) <- "Panic easily"
var_label(data1_clean$X21) <- "Be full of ideas"
var_label(data1_clean$X22) <- "Avoid difficult reading material"
var_label(data1_clean$X23) <- "Carry the conversation to a higher level"
var_label(data1_clean$X24) <- "Spend time reflecting on things"
var_label(data1_clean$X25) <- "Will not probe deeply into a subject"
```

An exploratory factor analysis (EFA) was conducted on the 25 personality items measured on a 6-point Likert scale. The dataset contained some missing value(NA), which
are handled using deletion. In odrer to determine the number of factors, two methods(PCA vs FA) were compared to improve explicity. 

```{r Q1-2, echo=knitr::is_html_output(), include=TRUE}
summary(data1_clean)

data1.pr <- prcomp(data1_clean, rotate="none")
summary(data1.pr)
pr.var = data1.pr$sdev^2
pve=pr.var/sum(pr.var)

plot(
  pve,
  xlab = "principal Component",
  ylab = "Proportion of Variance Explained",
  ylim = c(0, 1),
  type = "b"
)

plot(
  cumsum(pve),
  xlab = "principal Component",
  ylab = "Cummulative Proportion of Variance Explained",
  ylim = c(0, 1),
  type = "b"
)
cumsum(pve)
en <- eigen(cor(data1_clean))
en$values
plot(en$values, type = "o")


fa.parallel(
  data1_clean,
  fa = "fa",
  main = "Factor Analysis"
)



result <- factanal(data1_clean, factors = 6, rotation="varimax", scores="regression")

result


print(result, digits=2, cutoff=0.5)
print(result$loadings, cutoff=0)

result2 <- factanal(data1_clean, factors = 5, rotation="varimax", scores="regression")
result2

print(result2, digits=2, cutoff=0.5)
print(result2$loadings, cutoff=0)

```

# copula analysis 

```{r Q2, echo=knitr::is_html_output(), include=TRUE}
# library(GJRM)
# library(insuranceData)

data(dataCar)

data_cop <- subset(dataCar,
                   numclaims > 0 & claimcst0 > 0 & 
			 is.finite(claimcst0) & is.finite(exposure) & 
	   	   	 exposure > 0)

eq1 <- numclaims ~ 1
eq2 <- claimcst0 ~ 1

#claim frequency distribution
#use offset since exposure is not constant
#truncated frequency

tPfreq <- GJRM::gamlss(list(eq1), family = "tP", data = data_cop, offset = data_cop$exposure)
tNBIfreq <- GJRM::gamlss(list(eq1), family = "tNBI", data = data_cop, offset = data_cop$exposure)
tNBIIfreq <- GJRM::gamlss(list(eq1), family = "tNBII", data = data_cop, offset = data_cop$exposure)
tPIGfreq <- GJRM::gamlss(list(eq1), family = "tPIG", data = data_cop, offset = data_cop$exposure) 

print(AIC(tPfreq, tNBIfreq, tNBIIfreq, tPIGfreq))
print(BIC(tPfreq, tNBIfreq, tNBIIfreq, tPIGfreq))
```


```{r}
conv.check(tPfreq)
res.check(tPfreq)
summary(tPfreq)
```


TODO: Marcella put her analysis results here

it's using markdown syntax.

Example latex:
$$
x+1
$$


```{r}

#claim severity distribution

LNsev <- GJRM::gamlss(list(eq2), family="LN", data=data_cop)
IGsev <- GJRM::gamlss(list(eq2), family="IG", data=data_cop)
GAsev <- GJRM::gamlss(list(eq2), family="GA", data=data_cop)

WEIsev <- GJRM::gamlss(list(eq2), family="WEI", data=data_cop)
SMsev <- GJRM::gamlss(list(eq2), family="SM", data=data_cop)
FISKsev <- GJRM::gamlss(list(eq2), family="FISK", data=data_cop)
DAGUMsev <- GJRM::gamlss(list(eq2), family="DAGUM", data=data_cop)

print(AIC(DAGUMsev, LNsev, WEIsev, IGsev, GAsev, SMsev, FISKsev))
print(BIC(DAGUMsev, LNsev, WEIsev, IGsev, GAsev, SMsev, FISKsev))
```


```{r}

conv.check(IGsev)
res.check(IGsev)
summary(IGsev)
```


```{r}

###fitting copula
ZTPIGN <- gjrm(list(eq1, eq2, ~1, ~1), data = data_cop, margin = c("tP", "IG"), model = "B", copula = "N")
ZTPIGT <- gjrm(list(eq1, eq2, ~1, ~1), data = data_cop, margin = c("tP", "IG"), model = "B", copula = "T")
```


```{r}
print(AIC(ZTPIGN, ZTPIGT))
```

```{=latex}
\newpage
```

# principal component Analysis

We applied Principal Component Analysis (PCA) to the dataset consisting of 11 variables. Because the variables are measured on different scales (e.g., income levels vs. rating scores), we standardised all variables using \(\texttt{scale.=TRUE}\). Missing values were removed using \(\texttt{na.omit}\).

```{r Q3 data, echo=knitr::is_html_output(), include=FALSE}
## R Markdown

data3 <- read.table("dataset3.txt", header = TRUE)
data3 <- na.omit(data3)
colnames(data3) <- c(
  "INC",
  "AGE",
  "SPEND",
  "SHOP",
  "LOYAL",
  "WEB",
  "SOCIAL",
  "EMAIL",
  "SAT",
  "PROMO",
  "CART"
)

# PCA with standardisation
pr.out <- prcomp(data3, scale.  = TRUE)

# variance explained
pr.var <- pr.out$sdev^2
pve <- pr.var / sum(pr.var)

```

The correlation matrix shows generally weak to moderate correlations among variables, suggesting that several variables related to purchasing behaviour (spending score, Shopping Frequency, and Brand loyalty) are strongly positively correlated. In contrast, Age shows a negative relationship with these variables
This indicates that the first principal component is likely to capture a consumer spending behaviour dimension


```{r Q3Cor, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Correlation matrix (numbers left, heatmap right)", fig.show='hold', fig.align='center', fig.width=6, fig.height=3.2, fig.pos='H', out.width='45%'}
      
corrplot(cor(data3),
         method="number",
         type="upper",
         tl.cex = 0.25, number.cex = 0.65, tl.srt = 45, cl.cex=0.45)

corrplot(cor(data3),
         method="color",
         type="upper",
         col=colorRampPalette(c("blue","white","red"))(200),
         tl.cex = 0.25, number.cex = 0.4, tl.srt = 45, cl.cex=0.45)

```


```{r Q3 PLOT1, eval=knitr::is_html_output(), echo=FALSE}

ggplot(data.frame(PC = 1:length(pve), PVE = pve), aes(x = PC, y = PVE)) +
  geom_line(color = "red", size = 1) +
  geom_point(size = 3, color = "darkgreen") +
  geom_text(
    aes(label = round(PVE, 2)),
    vjust = -0.8,
    color = "black",
    size = 4
  ) +
  ylim(0, 1) +
  labs(title = "Scree Plot with PVE", x = "Principal Component", y = "Proportion of Variance Explained") +
  theme_minimal()

```

The cumulative variance plot shows that the first few principal components explain a substantial proportion of the total variance. In particular, the first three components explain more than half of the overall variation. The parallel analysis also indicates that three principal components should be retained, as the first three observed eigenvalues exceed those obtained from simulated data. The loading plot suggests that PC1 captures purchasing behaviour variables such as Spending Score, Shopping Frequency, Brand Loyalty, and Income. PC2 is mainly associated with Age and Email Open Rate, while PC3 is largely driven by Social Media Engagement. Therefore, the dimensionality of the dataset can be effectively reduced to three principal components while retaining most of the relevant information.

```{r Q3plot2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Cumulative variance, parallel analysis.", fig.show='hold', fig.align='center', fig.width=7, fig.height=3, fig.pos='H', out.width='48%'}      
      
cum_pve <- cumsum(pve)

ggplot(data.frame(PC = 1:length(cum_pve), CUM = cum_pve), aes(x = PC, y = CUM)) +
  geom_line(color = "red", size = 0.6) +
  geom_point(size = 1.6, color = "darkgreen") +
  geom_hline(yintercept = 0.6,
             linetype = "dashed",
             color = "red") +
  geom_text(
    aes(label = round(CUM, 2)),
    vjust = -0.6,
    color = "black",
    size = 2.4
  ) +
  ylim(0, 1.05) +
  labs(title = "Cumulative Variance Explained", x = "Principal Component", y = "Cumulative PVE") +
  theme_minimal(base_size = 8)

op <- par(cex=0.5, mar=c(3,3,2,1))
fa.parallel(data3, fa="pc", main="Parallel analysis", n.iter=100)
par(op)

```


```{r Q3plot3, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="PCA loadings.", fig.show='hold', fig.align='center', fig.width=7, fig.height=3, fig.pos='H', out.width='48%'}      

load <- pr.out$rotation[, 1:3]
df <- melt(load)

ggplot(df, aes(Var2, Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white") +
  labs(title = "Loadings Heatmap (PC1~PC3)",
       x = "Principal Components", y = "Variables") +
  theme_minimal()

load2 <- as.data.frame(pr.out$rotation[,1:2])
load2$var <- rownames(load2)

ggplot(load2, aes(x=PC1, y=PC2, label=var)) +
  geom_hline(yintercept=0, linewidth=0.3) +
  geom_vline(xintercept=0, linewidth=0.3) +
  geom_segment(aes(x=0, y=0, xend=PC1, yend=PC2),
               linewidth=0.4,
               arrow = arrow(length = unit(0.08,"inches"))) +
  geom_text(size=2.4, vjust=-0.4) +
  coord_equal() +
  labs(title="Loadings plot",
       x="PC1 loading", y="PC2 loading") +
  theme_minimal(base_size=8)

```



```{r, echo=knitr::is_html_output(), include=TRUE}

km <- kmeans(data3, centers = 3)
scores <- data.frame(pr.out$x)
scores2 <- data.frame(pr.out$x[,1:2], cluster = factor(km$cluster))
scores3 <- data.frame(pr.out$x[, 1:3])  

round(pr.out$rotation[,1:3], 2)


 ggplot(scores2, aes(PC1, PC2, color = cluster)) +
   geom_point() +
   theme_minimal()

ggplot(df, aes(Var2, Var1, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low="blue", high="red", mid="white") +
  labs(title="PCA Loadings", x="PC", y="Variables") +
  theme_minimal(base_size=8)

ggplot(scores, aes(x = PC1, y = PC2)) +
  geom_point(color = "darkgreen", alpha = 0.7) +
  labs(title = "PCA Score Plot (PC1 vs PC2)", x = "PC1", y = "PC2") +
  theme_minimal()



plot_ly(scores3,
        x = ~PC1,
        y = ~PC2,
        z = ~PC3,
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 3, color = scores3$PC1, colorscale = "Viridis")) %>%
  layout(title = "3D PCA Plot (PC1, PC2, PC3)")

```


# Time seires
